---
title: "Dose REsponse models for bAyesian Model avERaging (dreamer)"
author: "Richard Payne"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dreamer}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## dreamer

The dreamer R package provides functions to easily perform Bayesian
model averaging on (longitudinal) dose-response models.  It also provides
functions to generate data, and some useful post-processing functions
including posterior means and quantiles, plots, and other metrics
(e.g. calculating the posterior probability minimum efficacious dose,
etc.).

See the "dreamer_method" vignette for a high-level overview of Bayesian model averaging and/or read Gould (2019) for the approach used by dreamer.

## Generate data
We can easily see the possible models to generate data from by typing
`?dreamer_data` in the console.  We will start by generating data
from a quadratic model.

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

```{r, message = FALSE}
library(dreamer)
library(dplyr)
library(ggplot2)
set.seed(888)
data <- dreamer_data_quad(
  n_cohorts = c(20, 20, 20, 20, 20), # number of subjects in each cohort
  doses = c(0, .5, 1, 1.5, 2), # dose administered to each cohort
  b1 = 5, # intercept
  b2 = 5,
  b3 = - 2,
  sigma = 2 # standard deviation
)
head(data)
ggplot(data, aes(dose, response)) + geom_point()
```

## Fit Model
Now we can use a Bayesian model averaging technique to analyze the data.
We will use a linear and an EMAX model as candidate models, each with prior
probability 1 / 2.  There are quite a number of candidate models including
quadratic, 

Hyperparameters for each model are set by using the `model_XXX()` functions.
Type `?model` for a description of the model constructors and the models
each of them fit.

```{r}
# BAYESIAN MODEL AVERAGING
output <- dreamer_mcmc(
  data = data,
  n_adapt = 1e3,
  n_burn = 1e3,
  n_iter = 1e4,
  n_chains = 2,
  silent = TRUE,
  # the argument name "mod_quad" is arbitrary and is chosen by the user
  mod_quad = model_quad(
    mu_b1 = 0,
    sigma_b1 = 10,
    mu_b2 = 0,
    sigma_b2 = 10,
    mu_b3 = 0,
    sigma_b3 = 10,
    shape = 1,
    rate = .001,
    w_prior = 1 / 2
  ),
  # the argument name "mod_emax" is arbitrary and is chosen by the user
  mod_emax = model_emax(
    mu_b1 = 0,
    sigma_b1 = 10,
    mu_b2 = 0,
    sigma_b2 = 10,
    mu_b3 = 0,
    sigma_b3 = 2,
    mu_b4 = 1,
    sigma_b4 = 5,
    shape = 1,
    rate = .001,
    w_prior = 1 / 2
  )
)
```

### Model Output

The output is a list of the sorted unique doses, unique times (for longitudinal
models), the prior and posterior weights
for each model, and MCMC output from each individual model.


```{r}
names(output)
```

Output can be easily summarized.  The `summary()` function provides posterior
estimates of parameters, Gelman diagnostics, and effective sample sizes for 
each model and parameter.

```{r}
summary(output)
```

### Posterior Credible Intervals

It is easy to find posterior credible intervals of the mean response
for doses of interest:

```{r}
posterior(output)

# posterior Pr(effect of dose - effect of reference dose)
posterior(x = output, reference_dose = 0) 
```

## Posterior Plots

Plot the Bayesian model averaging posterior over the dose range:

```{r}
plot(output)
```

One can also plot the observed means on top of the plot:

```{r}
plot(output, data = data)
```

One can also look the dose response relative to a specific dose:

```{r}
plot(output, reference_dose = 0) # adjust relative to dose = 0
```

The predictive credible intervals for the mean with `predictive` number
of observations can be added easily:

```{r}
plot(output, data = data, predictive = 10)
```

One could also look at the predictive distribution of the difference 
between two sample means from different doses:

```{r}
plot(output, reference_dose = 0, predictive = 10)
```

One can also plot individual model fits, if desired.  Here is the EMAX
model's posterior:

```{r}
plot(output$mod_emax, data = data)
```

Comparing all the model fits together is easy:

```{r}
plot_comparison(output)
```

If the plot is too crowded, you can select individual fits:

```{r}
plot_comparison(
  mod_emax = output$mod_emax,
  mod_quad = output$mod_quad
)
```

## Posterior Quantities
### Probability of Meeting Effect of Interest (EOI)

The posterior probability of meeting a certain effect of interest can be
calculated, i.e., Pr(mean response > EOI | data).
One can specify
custom doses and EOIs relative to another dose, or in absolute terms.

```{r}
# absolute: pr(mean at dose 0.50 > 1 | data)
pr_eoi(output, eoi = 1, dose = 0.50)

# relative: pr( (mean at dose 0.50) - (mean at dose 0.25) > 0.55 | data )
pr_eoi(output, eoi = 0.55, dose = 0.50, reference_dose = 0.25)

# vectorized
n_doses <- length(output$doses)
pr_eoi(
  output,
  eoi = rep(.55, n_doses),
  dose = output$doses,
  reference_dose = rep(0, n_doses)
)
```

Posterior probabilities of meeting the effect of interest can be calculated
for individual models by applying the `pr_eoi()` function to a specific model's
output:

```{r}
# from the quadratic model
pr_eoi(
  output$mod_quad,
  eoi = 1.5,
  dose = .75,
  reference_dose = 0
)
```

### Probability of Minimum Efficacious Dose (MED)

Calculating the probability each dose is the minimum efficacious dose
(as defined as the smallest dose which is above a certain clinically significant
difference (CSD)) in the
set of doses is performed with the `pr_med()` function:

```{r}
pr_med(output, csd = 8)

# placebo adjusted
pr_med(
  output,
  csd = 3,
  reference_dose = 0
)

# relative to placebo grp (dose = 0) for just the EMAX model
pr_med(
  output$mod_emax,
  csd = 3,
  reference_dose = 0
)
```

### Probability of Minimum Dose with at Least X% Efficacy

Calculating the posterior probability each of the specified doses is the
smallest dose with at least X% efficacy can be done as follows:

```{r}
# looking for smallest dose with 95% of the maximum efficacy
pr_medx(output, ed = 95)
```

### Posterior Probability of MEDX

These functions obtain posterior quantiles of the minimum efficacious dose
which has X% of maximum efficacy.

```{r}
post_medx(output, ed = 95)
```

### Posterior Distribution of Percentage Effect

This function provides the posterior quantiles of a dose's effective dose
percent.  This
is calculated by calculating the dose's response divided by the maximum
response over the range for each iteration of the MCMC.

```{r}
post_perc_effect(
  output,
  dose = c(.05, .5)
)
```

## MCMC Diagnostics

Get diagnostics for each parameter:

```{r}
diagnostics(output)
# single model
diagnostics(output$mod_emax)
```

MCMC traceplots are easy to plot:

```{r, fig.height = 8, fig.width = 6}
# traceplot for all parameters for each model
# plot_trace(output)

# single model
plot_trace(output$mod_quad)
```

## Visualizing Priors

To help users choose a prior, the `dreamer_plot_prior()` function allows the user
to see what the prior actually looks like in context of the dose range for 
a particular model.  This is particularly helpful for complex models, or
binary models.

```{r}
dreamer_plot_prior(
  doses = c(0, 2.5, 5),
  mod_linear_binary = model_linear_binary(
    mu_b1 = -1,
    sigma_b1 = .1,
    mu_b2 = 1,
    sigma_b2 = .1,
    link = "logit",
    w_prior = 1
  )
)
```

Individual prior draws can also be plotted:

```{r}
dreamer_plot_prior(
  doses = seq(from = 0, to = 5, length.out = 50),
  n_samples = 100,
  plot_draws = TRUE,
  mod_quad_binary = model_quad_binary(
    mu_b1 = -.5,
    sigma_b1 = .2,
    mu_b2 = -.5,
    sigma_b2 = .2,
    mu_b3 = .5,
    sigma_b3 = .1,
    link = "logit",
    w_prior = 1
  )
)
```

One can also view the prior for all the models combined in a Bayesian model
averaging prior!

```{r}
dreamer_plot_prior(
  doses = c(0, 2.5, 5),
  mod_linear_binary = model_linear_binary(
    mu_b1 = -1,
    sigma_b1 = .1,
    mu_b2 = 1,
    sigma_b2 = .1,
    link = "logit",
    w_prior = .75
  ),
  mod_quad_binary = model_quad_binary(
    mu_b1 = -.5,
    sigma_b1 = .2,
    mu_b2 = -.5,
    sigma_b2 = .2,
    mu_b3 = .5,
    sigma_b3 = .1,
    link = "logit",
    w_prior = .25
  )
)
```

## Independent Model
An independent mean model can also be fit to each of the doses.
In this case, because no parametric assumptions are made on the dose-response
curve, no Bayesian model averaging is employed and no interpolation is 
allowed.

```{r}
output_independent <- dreamer_mcmc(
  data = data,
  n_adapt = 1e3,
  n_burn = 1e3,
  n_iter = 1e4,
  n_chains = 2,
  silent = TRUE, # make rjags be quiet,
  # this model has the same prior on the mean for each dose
  mod_indep = model_independent(
    mu_b1 = 0,
    sigma_b1 = 1,
    shape = 1,
    rate = .001
  )
)
# prior is too strong!
plot(output_independent, data = data)

output_independent2 <- dreamer_mcmc(
  data = data,
  n_adapt = 1e3,
  n_burn = 1e3,
  n_iter = 1e4,
  n_chains = 2,
  silent = TRUE, # make rjags be quiet,
  # this model has the different priors on the mean for each dose
  mod_indep = model_independent(
    mu_b1 = c(0, 1, 2, 3, 4),
    sigma_b1 = c(10, 10, 20, 20, 30),
    shape = 1,
    rate = .001,
    doses = c(0, 0.5, 1, 1.5, 2)
  )
)
plot(output_independent2, data = data)
```

All other functions are also available as above.

## Longitudinal Modeling

Longitudinal modeling is also available in `dreamer`.  Longitudinal data
can be generated by specifying the `longitudinal` argument, the
longitudinal parameters, and the times to observe subjects.  See documentation
for `dreamer::model_longitudinal()` and `dreamer::model` for the
parameterization of the longitudinal models.

```{r}
set.seed(889)
data_long <- dreamer_data_linear(
  n_cohorts = c(10, 10, 10, 10), # number of subjects in each cohort
  doses = c(.25, .5, .75, 1.5), # dose administered to each cohort
  b1 = 0, # intercept
  b2 = 2, # slope
  sigma = .5, # standard deviation,
  longitudinal = "itp",
  times = c(0, 12, 24, 52),
  t_max = 52, # maximum time
  a = .5,
  c1 = .1
)

ggplot(data_long, aes(time, response, group = dose, color = factor(dose))) +
  geom_point()
```

Fitting the MCMC is the same as before, except priors need to be specified
for the longitudinal part of the model:

```{r}
# BAYESIAN MODEL AVERAGING
output_long <- dreamer_mcmc(
  data = data_long,
  n_adapt = 1e3,
  n_burn = 1e3,
  n_iter = 1e4,
  n_chains = 2,
  silent = TRUE, # make rjags be quiet,
  mod_linear = model_linear(
    mu_b1 = 0,
    sigma_b1 = 1,
    mu_b2 = 0,
    sigma_b2 = 1,
    shape = 1,
    rate = .001,
    w_prior = 1 / 2, # prior probability of the model
    longitudinal = model_longitudinal_itp(
      mu_a = 0,
      sigma_a = 1,
      a_c1 = 0,
      b_c1 = 1,
      t_max = 52
    )
  ),
  mod_quad = model_quad(
    mu_b1 = 0,
    sigma_b1 = 1,
    mu_b2 = 0,
    sigma_b2 = 1,
    mu_b3 = 0,
    sigma_b3 = 1,
    shape = 1,
    rate = .001,
    w_prior = 1 / 2,
    longitudinal = model_longitudinal_linear(
      mu_a = 0,
      sigma_a = 1,
      t_max = 52
    )
  )
)
```

Plotting longitudinal models is straightforward:

```{r}
plot(output_long, data = data_long)
# plot dose response at last time point
plot(output_long, times = 52, data = data_long)
```

All the other posterior quantity functions work on longitudinal data.
For example:

```{r}
posterior(output_long)
```

## Reference
Gould, A. Lawrence. "BMA‐Mod: A Bayesian model averaging strategy for determining dose‐response relationships in the presence of model uncertainty." *Biometrical Journal* 61.5 (2019): 1141-1159.